---
title: 'R code for "Tidal streams, fish and seabirds: understanding the linkages between
  mobile predators, prey and hydrodynamics"'
author: "Ana Couto"
date: "02/11/2021"
output: 
  html_document:
    theme: flatly
    highlight: monochrome
    code_folding: show
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: false
      smooth_scroll: false
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)
```

# Packages used

```{r packages}
library(mgcv)
library(mgcViz)
library(dplyr)
library(patchwork)
library(ggplot2)
library(gratia)
library(mapdata)
library(raster)
```

# Data manipulation

### Load 2016 data

```{r data2016}
setwd("F:/Scotia/Birds_paper/")
birds2016 <- read.csv("./data/spatial_birds_2016_turbulence_circular.csv")
birds2016$year <- 2016
```

#### Calculating lagged pelagic fish and sandeel presence for 2016

```{r fish2016}
#Calculate lagged fish and sandeels presence for 2016
sandeels <- data.frame(matrix(ncol = 3, nrow = 1))
pelagic_fish <- data.frame(matrix(ncol = 3, nrow = 1))
survey <- sort(unique(birds2016$survey))

for (r in 1:nrow(birds2016)) {
  s <- which(survey == birds2016$survey[r])
  l <- birds2016$loc_id[r]
  
  tryCatch( {sandeels[r, 1] <- birds2016$sandeels_presence[which(birds2016$survey == survey[s - 2] & birds2016$loc_id == l)]}, 
            error = function(e) {})
  
  tryCatch({sandeels[r, 2] <- birds2016$sandeels_presence[which(birds2016$survey == survey[s - 1] & birds2016$loc_id == l)]},
           error = function(e) {})
  
  sandeels[r, 3] <- birds2016$sandeels_presence[r]
  
  tryCatch({pelagic_fish[r, 1] <- birds2016$fish_presence[which(birds2016$survey == survey[s - 2] & birds2016$loc_id == l)]},
    error = function(e) {})
  
  tryCatch({ pelagic_fish[r, 2] <- birds2016$fish_presence[which(birds2016$survey == survey[s - 1] & birds2016$loc_id == l)]},
    error = function(e) {})
  
  pelagic_fish[r, 3] <- birds2016$fish_presence[r]
} #close for loop

lag_sandeels <- data.frame(matrix(ncol = 3, nrow = nrow(birds2016)))
for (r in 1:nrow(lag_sandeels)) {lag_sandeels[r, ] <- seq(-2, 0, 1)}

lag_sandeels <- as.matrix(lag_sandeels)

lag_pelagic_fish <- data.frame(matrix(ncol = 3, nrow = nrow(birds2016)))
for (r in 1:nrow(lag_pelagic_fish)) {
  lag_pelagic_fish[r, ] <- seq(-2, 0, 1)
}

lag_pelagic_fish <- as.matrix(lag_pelagic_fish)

pelagic_fish <- as.matrix(pelagic_fish)
pelagic_fish[is.na(pelagic_fish)] <- 0

sandeels <- as.matrix(sandeels)
sandeels[is.na(sandeels)] <- 0

birds2016$lag_pelagic_fish <- lag_pelagic_fish
birds2016$pelagic_fish <- pelagic_fish

birds2016$lag_sandeels <- lag_sandeels
birds2016$sandeels <- sandeels

rm(lag_sandeels)
rm(lag_pelagic_fish)
rm(sandeels)
rm(pelagic_fish)
```

### Load 2018 data

```{r data2018}
setwd("F:/Scotia/Birds_paper/")
birds2018 <- read.csv("./data/spatial_birds_2018_turbulence_circular.csv")
birds2018$year <- 2018
```

#### Calculating lagged pelagic fish and sandeel presence for 2018

```{r fish2018}
#Calculate lagged fish and sandeels presence for 2018
sandeels <- data.frame(matrix(ncol = 3, nrow = 1))
pelagic_fish <- data.frame(matrix(ncol = 3, nrow = 1))
survey <- sort(unique(birds2016$survey))

for (r in 1:nrow(birds2018)) {
  s <- which(survey == birds2018$survey[r])
  l <- birds2018$loc_id[r]
  
  tryCatch( {sandeels[r, 1] <- birds2016$sandeels_presence[which(birds2016$survey == survey[s - 2] & birds2016$loc_id == l)]}, 
            error = function(e) {})
  
  tryCatch({sandeels[r, 2] <- birds2016$sandeels_presence[which(birds2016$survey == survey[s - 1] & birds2016$loc_id == l)]},
           error = function(e) {})
  
  sandeels[r, 3] <- birds2016$sandeels_presence[r]
  
  
  tryCatch({pelagic_fish[r, 1] <- birds2018$fish_presence[which(birds2018$survey == survey[s - 2] & birds2018$loc_id == l)]},
    error = function(e) {})
  
  tryCatch({pelagic_fish[r, 2] <- birds2018$fish_presence[which(birds2018$survey == survey[s - 1] & birds2018$loc_id == l)]},
    error = function(e) {})
  
  pelagic_fish[r, 3] <- birds2018$fish_presence[r]
} #close for loop

lag_sandeels <- data.frame(matrix(ncol = 3, nrow = nrow(birds2018)))

for (r in 1:nrow(lag_sandeels)) {lag_sandeels[r, ] <- seq(-2, 0, 1)}

lag_sandeels <- as.matrix(lag_sandeels)

lag_pelagic_fish <- data.frame(matrix(ncol = 3, nrow = nrow(birds2018)))

for (r in 1:nrow(lag_pelagic_fish)) {lag_pelagic_fish[r, ] <- seq(-2, 0, 1)}

lag_pelagic_fish <- as.matrix(lag_pelagic_fish)

pelagic_fish <- as.matrix(pelagic_fish)
pelagic_fish[is.na(pelagic_fish)] <- 0

sandeels <- as.matrix(sandeels)
sandeels[is.na(sandeels)] <- 0

birds2018$lag_pelagic_fish <- lag_pelagic_fish
birds2018$pelagic_fish <- pelagic_fish

birds2018$lag_sandeels <- lag_sandeels
birds2018$sandeels <- sandeels
```

### Merge 2016 and 2018

```{r merge}
#Merge datasets and remove NA
full.data <- rbind(birds2016, birds2018)
full.data <- full.data[complete.cases(full.data), ]

#Transform section , Transect and year into factor
full.data$Section <- as.factor(full.data$loc_id)
full.data$Transect <- as.factor(paste(full.data$survey, full.data$year, sep="_"))
full.data$year <- as.factor(full.data$year)

# calculate offset
full.data$offset <- full.data$npoints

#select daytime only for seabirds
full.data<-full.data[which(full.data$hour %in% seq(4,19,1)),]
```

# Data visualisation

### Plot mean number of seabirds per section

```{r seabirds}
a <- ggplot(full.data, aes(x = loc_id, y =  benthic_feeder, group=year))+
  geom_rect(aes(xmin=23, xmax=24, ymin=-Inf, ymax=Inf), fill="lightgrey",  fill.alpha=0.5,inherit.aes = FALSE)+
  stat_summary(geom = "errorbar", fun.data = mean_se, alpha = 0.2) +
  stat_summary(geom = "line", fun = mean,  show.legend = T,aes(group=year, linetype=year), size=1) +
  geom_vline(xintercept = c(19,36), linetype = 2) +
  theme_classic() +
  labs(subtitle = "Benthic", 
       x = "Section",
       y = "Mean number of \nseabirds")

b <- ggplot(full.data, aes(x = loc_id, y =  pelagic_feeder, group=year))+ 
  geom_rect(aes(xmin=23, xmax=24, ymin=-Inf, ymax=Inf), fill="lightgrey",  fill.alpha=0.5,inherit.aes = FALSE)+
  stat_summary(geom = "errorbar", fun.data = mean_se, alpha = 0.2) +
  stat_summary(geom = "line", fun = mean,  show.legend = T,aes(group=year, linetype=year), size=1) +
  geom_vline(xintercept = c(19,36), linetype = 2) +
  theme_classic() +
  labs(subtitle = "Pelagic",
       x = "Section",
       y = "Mean number of \nseabirds")

c <- ggplot(full.data, aes(x = loc_id, y =  gu, group=year))+ 
  geom_rect(aes(xmin=23, xmax=24, ymin=-Inf, ymax=Inf), fill="lightgrey",  fill.alpha=0.5,inherit.aes = FALSE)+
  stat_summary(geom = "errorbar", fun.data = mean_se, alpha = 0.2) +
  stat_summary(geom = "line", fun = mean,  show.legend = T,aes(group=year, linetype=year), size=1) +
  geom_vline(xintercept = c(19,36), linetype = 2) +
  theme_classic() +
  labs(subtitle = "Common guillemots",
       x = "Section",
       y = "Mean number of \nseabirds")

nested_1 <- a + b +c 

seabirds<- nested_1 +
  plot_layout(ncol = 1) +
  plot_annotation(tag_levels = "A", tag_suffix = ")")

seabirds
```

### Plot mean environmental variables per section

```{r env}
d <- ggplot(full.data, aes(x = loc_id, y =  sandeels_presence, group=year)) +
  geom_rect(aes(xmin=23, xmax=24, ymin=-Inf, ymax=Inf), fill="lightgrey",  fill.alpha=0.5,inherit.aes = FALSE)+
  stat_summary(geom = "errorbar", fun.data = mean_se, alpha = 0.2) +
  stat_summary(geom = "line", fun = mean,  show.legend = T,aes(group=year, linetype=year), size=1) +
  geom_vline(xintercept = c(19,36), linetype = 2) +
  theme_classic() +
  labs(subtitle = "Sandeels", 
       x = "Section",
       y = "Mean schools presence")

e <- ggplot(full.data, aes(x = loc_id, y =  fish_presence, group=year)) +
  geom_rect(aes(xmin=23, xmax=24, ymin=-Inf, ymax=Inf), fill="lightgrey",  fill.alpha=0.5,inherit.aes = FALSE)+
  stat_summary(geom = "errorbar", fun.data = mean_se, alpha = 0.2) +
  stat_summary(geom = "line", fun = mean,  show.legend = T,aes(group=year, linetype=year), size=1) +
  geom_vline(xintercept = c(19,36), linetype = 2) +
  theme_classic() +
  labs(subtitle = "Non-sandeels", 
       x = "Section",
       y = "Mean school presence")

f <- ggplot(full.data, aes(x = loc_id, y =  velocity, group=year)) +
  geom_rect(aes(xmin=23, xmax=24, ymin=-Inf, ymax=Inf), fill="lightgrey",  fill.alpha=0.5,inherit.aes = FALSE)+
  stat_summary(geom = "errorbar", fun.data = mean_se, alpha = 0.2) +
  stat_summary(geom = "line", fun = mean,  show.legend = T,aes(group=year, linetype=year), size=1) +
  geom_vline(xintercept = c(19,36), linetype = 2) +
  theme_classic() +
  labs(
    x = "Section",
    y = "Velocity (m/s)")

g <- ggplot(full.data, aes(x = loc_id, y =  proportion, group=year)) +
  geom_rect(aes(xmin=23, xmax=24, ymin=-Inf, ymax=Inf), fill="lightgrey",  fill.alpha=0.5,inherit.aes = FALSE)+
  stat_summary(geom = "errorbar", fun.data = mean_se, alpha = 0.2) +
  stat_summary(geom = "line", fun = mean,  show.legend = T,aes(group=year, linetype=year), size=1) +
  geom_vline(xintercept = c(19,36), linetype = 2) +
  theme_classic() +
  labs(
    x = "Section",
    y = "Turbulence interference \nproportion (%)")

nested_2 <- d + e +f +g

env_variables<- nested_2 +
  plot_layout(ncol = 2) +
  plot_annotation(tag_levels = "A", tag_suffix = ")")

env_variables
#ggsave(file="env_variables_1618.jpg",env_variables, width =20, height = 15, units = "cm") #saves plot
```

# Statistical models

#### Remove data where turbulence proportion exceeds 10%
```{r 10%}
data10<-full.data[which(full.data$proportion<10), ]
data_rm<-full.data[which(full.data$proportion>=10), ]
data_rm$n<-1
summary_rm<-aggregate(cbind(fish_presence, sandeels_presence, n)~ loc_id+year, data_rm, sum)

```

#### GAMM for benthic seabirds
##### Run the model
```{r benthic}
benthic10<-gam(benthic_feeder ~ year+
                        s(Transect, bs="re") + # random effect 
                        s(Section, bs="re") + # random effect
                        s(velocity) +
                        s(lag_pelagic_fish, by= pelagic_fish, k= 3)+
                        s(lag_sandeels, by= sandeels, k= 3)+
                        offset(log(offset)), family= "nb",method= "REML", data= data10, select=T)
```

##### Model summary
```{r}
summary(benthic10)
```

##### Model appraisal 

```{r}
appraise(benthic10)
concurvity(benthic10)
```
##### Model figures


```{r}

#plot splines
b10 <- getViz(benthic10)

b1<-plot(sm(b10, select = 3),
            # To include se with mean
            seWithMean = TRUE,
            # 1000 takes me a while to plot
            nsim= 1000)+
  # generic backtransform code
  # layer to add in sims
  l_simLine() +
  l_fitLine(colour = "black", size = 1) +
  l_ciPoly(level = 0.95, fill = "orange", alpha = 0.4) +
  theme_classic()+geom_hline(yintercept = 0, lty=2) + labs(y = "Effect size", x="Velocity", caption = "k = 9, EDF =  0.85") 

b2<-plot(sm(b10, select = 4),
         # To include se with mean
         seWithMean = TRUE,
         # 1000 takes me a while to plot
         nsim= 1000)+
  # generic backtransform code
  # layer to add in sims
  l_simLine() +
  l_fitLine(colour = "black", size = 1) +
  l_ciPoly(level = 0.95, fill = "orange", alpha = 0.4) +
  theme_classic()+ geom_hline(yintercept = 0, lty=2) + labs(y = "Effect size", x="Non-sandeel schools presence", caption = "k = 3, EDF =  0.89") 


b3<-plot(sm(b10, select = 5),
         # To include se with mean
         seWithMean = TRUE,
         # 1000 takes me a while to plot
         nsim= 1000)+
  # generic backtransform code
  # layer to add in sims
  l_simLine() +
  l_fitLine(colour = "black", size = 1) +
  l_ciPoly(level = 0.95, fill = "orange", alpha = 0.4) +
  theme_classic()+ geom_hline(yintercept = 0, lty=2) + labs(y = "Effect size", x="Sandeel schools presence",caption = "k = 3, EDF =  0.9") 

#### merge all figures
nested_benthic<- b1$ggObj + b2$ggObj + b3$ggObj  
  model_fig_benthic<- nested_benthic + plot_annotation(tag_levels = "A") + plot_layout(ncol = 1)
model_fig_benthic
```

##### Run model without random effects

```{r}
benthic10_no_re<-gam(benthic_feeder ~ year+
                        s(velocity) +
                        s(lag_pelagic_fish, by= pelagic_fish, k= 3)+
                        s(lag_sandeels, by= sandeels, k= 3)+
                        offset(log(offset)), sp= benthic10$sp[3:8], family= "nb",method= "REML", data= data10, select=T)
summary(benthic10_no_re)
```


#### GAMM for pelagic seabirds
##### Run the model

```{r}
pelagic10<-gam(pelagic_feeder ~ year+
                 s(Transect, bs="re") + # random effect 
                 s(Section, bs="re") + # random effect
                 s(velocity) +
                 s(lag_pelagic_fish, by= pelagic_fish, k= 3)+
                 s(lag_sandeels, by= sandeels, k= 3)+
                 offset(log(offset)), family= "nb",method= "REML", data= data10, select=T)
```

##### Model summary

```{r}
summary(pelagic10)
```

##### Model appraisal

```{r}
appraise(pelagic10)
concurvity(pelagic10)
```

##### Model figures

```{r}
p10 <- getViz(pelagic10)
p1<-plot(sm(p10, select = 3),
         # To include se with mean
         seWithMean = TRUE,
         # 1000 takes me a while to plot
         nsim= 1000)+
  # generic backtransform code
  # layer to add in sims
  l_simLine() +
  l_fitLine(colour = "black", size = 1) +
  l_ciPoly(level = 0.95, fill = "orange", alpha = 0.4) +
  theme_classic()+geom_hline(yintercept = 0, lty=2) + labs(y = "Effect size", x="Velocity", caption = "k = 9, EDF = 3.38") 

p2<-plot(sm(p10, select = 4),
         # To include se with mean
         seWithMean = TRUE,
         # 1000 takes me a while to plot
         nsim= 1000)+
  # generic backtransform code
  # layer to add in sims
  l_simLine() +
  l_fitLine(colour = "black", size = 1) +
  l_ciPoly(level = 0.95, fill = "orange", alpha = 0.4) +
  theme_classic()+ geom_hline(yintercept = 0, lty=2) + labs(y = "Effect size", x="Non-sandeel schools presence",caption = "k = 3, EDF =  1.83") 

p3<-plot(sm(p10, select = 5),
         # To include se with mean
         seWithMean = TRUE,
         # 1000 takes me a while to plot
         nsim= 1000)+
  # generic backtransform code
  # layer to add in sims
  l_simLine() +
  l_fitLine(colour = "black", size = 1) +
  l_ciPoly(level = 0.95, fill = "orange", alpha = 0.4) +
  theme_classic()+ geom_hline(yintercept = 0, lty=2) + labs(y = "Effect size", x="Sandeel schools presence",caption = "k = 3, EDF =  0.29") 

#### merge all figures
nested_pelagic<- p1$ggObj + p2$ggObj + p3$ggObj  
model_fig_pelagic<- nested_pelagic +plot_annotation(tag_levels = "A") + plot_layout(ncol = 1)
model_fig_pelagic
```

##### Run model without random effects

```{r}
pelagic10_no_re<-gam(pelagic_feeder ~ year+
                        s(velocity) +
                        s(lag_pelagic_fish, by= pelagic_fish, k= 3)+
                        s(lag_sandeels, by= sandeels, k= 3)+
                        offset(log(offset)), sp= pelagic10$sp[3:8], family= "nb",method= "REML", data= data10, select=T)
summary(pelagic10_no_re)
```

#### GAMM for common guillemots
##### Run the model

```{r}
guillemots10<-gam(gu ~ year+
                    s(Transect, bs="re") + # random effect 
                    s(Section, bs="re") + # random effect
                    s(velocity) +
                    s(lag_pelagic_fish, by= pelagic_fish, k= 3)+
                    s(lag_sandeels, by= sandeels, k= 3)+
                    offset(log(offset)), family= "nb",method= "REML", data= data10, select=T)
```

##### Model summary

```{r}
summary(guillemots10)
```

##### Model appraisal

```{r}
appraise(guillemots10)
```

##### Model figures

```{r}
g10 <- getViz(guillemots10)
g1<-plot(sm(g10, select = 3),
         # To include se with mean
         seWithMean = TRUE,
         # 1000 takes me a while to plot
         nsim= 1000)+
  # generic backtransform code
  # layer to add in sims
  l_simLine() +
  l_fitLine(colour = "black", size = 1) +
  l_ciPoly(level = 0.95, fill = "orange", alpha = 0.4) +
  theme_classic()+geom_hline(yintercept = 0, lty=2) + labs(y = "Effect size", x="Velocity", subtitle="Velocity", caption = "k = 9, EDF = 5.6") 

g2<-plot(sm(g10, select = 4),
         # To include se with mean
         seWithMean = TRUE,
         # 1000 takes me a while to plot
         nsim= 1000)+
  # generic backtransform code
  # layer to add in sims
  l_simLine() +
  l_fitLine(colour = "black", size = 1) +
  l_ciPoly(level = 0.95, fill = "orange", alpha = 0.4) +
  theme_classic()+ geom_hline(yintercept = 0, lty=2) + labs(y = "Effect size", x="Non-sandeel schools presence", caption = "k = 3, EDF = 1.94") 

g3<-plot(sm(g10, select = 5),
         # To include se with mean
         seWithMean = TRUE,
         # 1000 takes me a while to plot
         nsim= 1000)+
  # generic backtransform code
  # layer to add in sims
  l_simLine() +
  l_fitLine(colour = "black", size = 1) +
  l_ciPoly(level = 0.95, fill = "orange", alpha = 0.4) +
  theme_classic()+ geom_hline(yintercept = 0, lty=2) + labs(y = "Effect size", x="Sandeel schools presence",caption = "k = 3, EDF =  1.23") 

#### merge all figures
nested_guillemots<- g1$ggObj + g2$ggObj + g3$ggObj 
model_fig_guillemots<- nested_guillemots +plot_annotation(tag_levels = "A") + plot_layout(ncol = 1)
model_fig_guillemots
```

##### Run model without random effects

```{r}
guillemots10_no_re<-gam(gu ~ year+
                        s(velocity) +
                        s(lag_pelagic_fish, by= pelagic_fish, k= 3)+
                        s(lag_sandeels, by= sandeels, k= 3)+
                        offset(log(offset)), sp= guillemots10$sp[3:8], family= "nb",method= "REML", data= data10, select=T)
summary(guillemots10_no_re)
```

